---
title: "SLE777_Renee Carroll_Assessment 4"
author: "Renee Carroll"
date: "2025-09-25"
output: 
  html_document:
    toc: yes
theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1
## Gene Expression Data

The file “gene_expression.tsv” contains RNA-seq count data for three samples of interest.

# 1. Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes.

```{r download files from GitHub, and show first 6 genes}
URL1 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL1, destfile="gene_expression.tsv")
gene_expression <- read.delim("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t", stringsAsFactors = FALSE)
head(gene_expression)
```

# 2. Make a new column which is the mean of the other columns. Show a table of values for the first six genes.

```{r mean saved as new column}
gene_expression$mean_expression <- rowMeans(gene_expression)
head(gene_expression)
```

# 3. List the 10 genes with the highest mean expression

```{r 10 genes with highest mean - whole table}
ordered_expression <- gene_expression[order(-gene_expression$mean_expression),]
genes_highest_expression <- head(ordered_expression, 10)
genes_highest_expression
```

# Listing just the 10 gene names

```{r 10 genes with highest mean - just names as a list}
rownames(genes_highest_expression)
```

# 4. Determine the number of genes with a mean < 10

```{r number of genes with low expression < 10}
low_exp_genes <- subset(gene_expression, mean_expression<10)
nrow(low_exp_genes)
```

# 5. Make a histogram plot of the mean values and include it into your report.

```{r histogram, fig_width: 7, fig_height: 7}
hist(gene_expression$mean_expression, breaks=1000, main = "Mean Gene Expression", xlab="Mean Gene Expression", col="dodgerblue2")
```

As the data is highly positively skewed, it is best represented on a logarithmic scale.

```{r}
log_gene_expression <- log(gene_expression+1) # Add 1 to each data point, to account for zeros in the dataset, as you cant take a log of zero.
hist(log_gene_expression$mean_expression, breaks=50, main = "Mean Gene Expression", xlab="Log Mean Gene Expression", col="dodgerblue2")
```
As the graph is still highly skewed, lets remove lowly expressed genes with mean expression less than 10.

```{r histogram of log of values}
gene_subset_expressionhigherthan10 <- subset(gene_expression, mean_expression>10)
log_gene_subset_expressionhigherthan10 <- log(gene_subset_expressionhigherthan10)
hist(log_gene_subset_expressionhigherthan10$mean_expression, breaks=50, main = "Mean Gene Expression", xlab = "Log Mean Gene Expression", col = "dodgerblue2")
```
## Part 1
## Tree Circumference Data

The file “growth_data.csv” contains measurements for tree circumference growing at two sites, control site and treatment site which were planted 20 years ago.

# 6. Import this csv file into an R object. What are the column names?

```{r Import csv file}
URL2 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
download.file(URL2, destfile="growth_data.csv")
growth_data <- read.csv("growth_data.csv") # read.csv automatically treats the first row as headings, therefore you don't need to specify it in the code
```

```{r determine column names}
colnames(growth_data)
```

# 7. Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.

```{r}
girth_stats_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  Mean_2005 = c(mean(subset(growth_data, Site == "northeast")$Circumf_2005_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  SD_2005 = c(sd(subset(growth_data, Site == "northeast")$Circumf_2005_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  Mean_2020 = c(mean(subset(growth_data, Site == "northeast")$Circumf_2020_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2020_cm)),
  SD_2020 = c(sd(subset(growth_data, Site == "northeast")$Circumf_2020_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2020_cm))
)
print(girth_stats_df)
```

# 8. Make a box plot of tree circumference at the start and end of the study at both sites.

```{r boxplot of tree circumference at the start and end of the study at both sites}
boxplot(
  subset(growth_data, Site == "northeast")$Circumf_2005_cm, 
  subset(growth_data, Site == "northeast")$Circumf_2020_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2005_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2020_cm, 
  names = c("Northeast 2005", "Northeast 2020", "Southwest 2005", "Southwest 2020"), 
  ylab = "Circumference (cm)", 
  main = "Increase in tree girth between 2005 and 2020, at each study site")
```

# 9. Calculate the mean growth over the last 10 years at each site.

```{r mean growth over the last 10 years, northeast site}
# Add a column to the existing dataframe, to contain the calculation of 2020 measurements minus the 2010 measurements.
growth_data$growth_2010_2020 <- growth_data$Circumf_2020_cm-growth_data$Circumf_2010_cm 

# Create a new data frame to present the mean of the 10 year growth data, for each location.
mean_10yr_growth_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  "Mean 10yr growth" = c(mean(subset(growth_data, Site == "northeast")$growth_2010_2020), mean(subset(growth_data, Site == "southwest")$growth_2010_2020))
)
print(mean_10yr_growth_df)
```

# 10. Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.
The t.test() function returns an object containing various information, including:
t-statistic: The calculated t-value.
df: Degrees of freedom.
p-value: The probability of observing a t-statistic as extreme as, or more extreme than, the calculated one, assuming the null hypothesis is true. A small p-value (typically < 0.05) suggests evidence against the null hypothesis.
confidence interval: A range within which the true population mean difference (for two-sample tests) or population mean (for one-sample tests) is likely to fall.
sample means: The means of the groups or sample.

Assumptions:
T-tests rely on assumptions like normality of data (or residuals for two-sample tests) and, for independent samples, potentially equality of variances. You may need to check these assumptions before interpreting the results.

```{r test for normality in the northeast data}
hist(subset(growth_data, Site == "northeast")$growth_2010_2020, breaks = 20)
```

```{r test for normality in the southwest data}
hist(subset(growth_data, Site == "southwest")$growth_2010_2020, breaks = 20)
```


```{r t.test, not assuming equal variance (Welch t.test)}
t.test(growth_2010_2020 ~ Site, data = growth_data, var.equal = FALSE) #var.equal = TRUE: Include this if you assume equal variances between the groups (for a Student's t-test). Omit or set to FALSE for Welch's t-test (which doesn't assume equal variances).
```
```{r t.test, assuming equal variance}
t.test(growth_2010_2020 ~ Site, data = growth_data, var.equal = TRUE) #var.equal = TRUE: Include this if you assume equal variances between the groups (for a Student's t-test). Omit or set to FALSE for Welch's t-test (which doesn't assume equal variances).
```

## Part 2 Examining biological sequence diversity

In order to analyse sequence data, there are two packages we need to first install and load.

Seqinr is a package designed to process and analyse sequence data.
R.utils is a package containing general utilities like zip and unzip

```{r installing packages}
#install.packages("seqinr")
#install.packages("R.utils")
```


```{r loading packages}
suppressPackageStartupMessages({
  library("seqinr")
  library("R.utils")
})
```

# 1. Download the whole set of coding DNA sequences for E. coli and your organism of interest. How many coding sequences are present in these organisms? Present this in the form of a table. Describe any differences between the two organisms. 

```{r Download E.coli genome}
URL3 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_0_collection/escherichia_coli_str_k_12_substr_mg1655_gca_000005845/cds/Escherichia_coli_str_k_12_substr_mg1655_gca_000005845.ASM584v2.cds.all.fa.gz"
download.file(URL3, destfile="ecoli_cds.fa.gz")
gunzip("ecoli_cds.fa.gz")
```

```{r Download organism of interest, Acetobacter aceti GCA_002005445}
URL4 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_80_collection/acetobacter_aceti_gca_002005445/cds/Acetobacter_aceti_gca_002005445.ASM200544v1.cds.all.fa.gz"
download.file(URL4, destfile="aaceti_cds.fa.gz")
gunzip("aaceti_cds.fa.gz")
```
```{r reading in the two sequence data files}
ecoli_cds <- seqinr::read.fasta("ecoli_cds.fa")
aaceti_cds <- seqinr::read.fasta("aaceti_cds.fa")
```

```{r}
str(head(ecoli_cds))
```

```{r}
str(head(aaceti_cds))
```

```{r Table of the number of coding sequences present in these organisms}
cds_df <- data.frame(
  Organism = c("E.coli", "A.aceti"),
  "Number of Coding Sequences" = c(length(ecoli_cds), length(aaceti_cds)),
  check.names = FALSE 
  )
print(cds_df)
```

# Describe any differences between the two organisms. 
```{r percentage increase in number of coding sequences}
percentage_increase <- (length(ecoli_cds)-length(aaceti_cds))/length(aaceti_cds)*100
print(percentage_increase)
```
The organism E.coli contains 24.6% more genes than the organism A.aceti, suggesting E.coli is a more complex organism.

# 2. How much coding DNA is there in total for these two organisms? Present this in the form of a table. Describe any differences between the two organisms.
```{r Table of the total amount of coding DNA for the two organisms}
length_ecoli_genes <- as.numeric(summary(ecoli_cds)[,1])
length_aaceti_genes <- as.numeric(summary(aaceti_cds)[,1])
cds_total_df <- data.frame(
  Organism = c("E.coli", "A.aceti"), 
  "Total Amount of Coding Sequence" = c(sum(length_ecoli_genes), sum(length_aaceti_genes)),
  check.names = FALSE 
) # Including the check.names = FALSE code so that R doesn't add '.' in between the words of the Column names.
print(cds_total_df)
```
# Describe any differences between the two organisms. 
```{r percentage increase in total amount of coding sequences}
percentage_increase_total <- (sum(length_ecoli_genes)-sum(length_aaceti_genes))/sum(length_aaceti_genes)*100
print(percentage_increase_total)
```
In line with a 24.6% increase in genes, E.coli contains a proportional increase in total coding sequence, with an increase of 22.9% above the total amount of coding DNA contained in A.aceti organisms.


# 3. Calculate the length of all coding sequences in these two organisms. Make a boxplot of coding sequence length in these organisms. What is the mean and median coding sequence length of these two organisms? Describe any differences between the two organisms.
```{r Boxplot of coding sequence length in these organisms}
boxplot(length_ecoli_genes, length_aaceti_genes, names = c("E.coli", "A.aceti"), main = "Coding DNA Sequence Length in E.coli and A.aceti", ylab="Length (bp)")
```
```{r Boxplot of log coding sequence length in these organisms}
boxplot(log(length_ecoli_genes), log(length_aaceti_genes), names = c("E.coli", "A.aceti"), main = "Coding DNA Sequence Length in E.coli and A.aceti", ylab="Log of Length (bp)")
```
```{r Density plot to explore coding cDNA lengths between species}
# Create the vectors of the coding sequences
d_ecoli <- density(length_ecoli_genes)
d_aaceti <- density(length_aaceti_genes)

# Create the density plot
plot(d_ecoli, col="red", main = "Density plot of coding DNA sequence lengths", xlab = "cDNA length")
lines(d_aaceti, col="blue")

# Add custom tick marks
axis(1, at = seq(0, max(length_ecoli_genes, length_aaceti_genes), by = 1000))

# Add names for each line
text(x = 6000, y = 6e-04, labels = "E.coli", col = "red", cex = 1.2)
text(x = 6000, y = 5e-04, labels = "A.aceti", col = "blue", cex = 1.2)
```


```{r Mean and median coding sequence length in these organisms}
cds_stats_df <- data.frame(
  Organism = c("E.coli", "A.aceti"),
  "Mean cDNA Sequence Length" = c(mean(length_ecoli_genes), mean(length_aaceti_genes)),
  "Median cDNA Sequence Length" = c(median(length_ecoli_genes), median(length_aaceti_genes)),
  check.names = FALSE 
)
print(cds_stats_df)
```

# Describe any differences between the two organisms. 






# 4. Calculate the frequency of DNA bases in the total coding sequences for both organisms. Perform the same calculation for the total protein sequence. Create bar plots for nucleotide and amino acid frequency. Describe any differences between the two organisms.

```{r Nucleotide composition of E.coli}
# Unlist the DNA sequences from the 'ecoli_cds' data frame, save the sequences as a single vector called 'dna_ecoli'
dna_ecoli <- unlist(ecoli_cds) 

# Perform this code to confirm this vectors length is the same as the total number of nucleotides as determined earlier for E.coli in Step 2 above, ie 3978528. 
summary(dna_ecoli)

# Count the number of each nucleotide in the vector 'dna_ecoli', in combinations of 1's (ie. single nucleotides), save the sequences as a single vector called 'dna_composition_ecoli'
dna_composition_ecoli <- count(dna_ecoli,1)

# This will show the total number of each base in E.coli's coding sequence.
print(dna_composition_ecoli)

# Calculate the proportion of each nucleotide
dna_proportion_ecoli <- dna_composition_ecoli/sum(dna_composition_ecoli)
dna_proportion_ecoli <- round(dna_proportion_ecoli, digits = 3)
print(dna_proportion_ecoli)
```

```{r Nucleotide composition of A.aceti}
# As per the code for E.coli, now do the same for A.aceti
dna_aaceti <- unlist(aaceti_cds) 
summary(dna_aaceti)
dna_composition_aaceti <- count(dna_aaceti,1) 
print(dna_composition_aaceti)
dna_proportion_aaceti <- dna_composition_aaceti/sum(dna_composition_aaceti)
dna_proportion_aaceti <- round(dna_proportion_aaceti, digits = 3)
print(dna_proportion_aaceti)
```
```{r translating the coding sequences}
# Translate the cDNA from E.coli. The lapply function performs a loop, applying the 'translate' code to all elements of the 'ecoli_cds' data, and saving the protein sequences to the 'ecoli_prot' object.
ecoli_prot <- lapply(ecoli_cds, translate)

#translate the cDNA from A.aceti
aaceti_prot <- lapply(aaceti_cds, translate)
```

```{r}
head(ecoli_prot)
```

you see that this vector has a length of 1326176 characters, which is the total number of nucleotides (3978528, as determined earlier for E.coli in Step 2 above) divided by 3, as expected. 

```{r Determining the amino acid composition of E.coli proteome}
# Unlist the protein sequences from the 'ecoli_prot' data frame, saves the sequences as a single vector called 'ecoli_prot_total'
ecoli_prot_total <- unlist(ecoli_prot)

# Show the summary of the new vector, ecoli_prot_total, to see the total number of characters. This should be the total number of coding nucleotides divided by 3.
summary(ecoli_prot_total)

# Creates the list of possible amino acids to form our alphabet to use in the 'count' code.
aa <- unique(ecoli_prot_total) 

# Removes the stop codon asterisks.
aa <- aa[aa != "*"] 

# Confirm there are 20 amino acids, and what their identifyers are
length(aa)
print(aa)

# this counts the number of each amino acids in the vector 'aa_composition_ecoli', in combinations of 1's (ie. single amino acids), using the aa alphabet we just determined, and saves it as the vector 'aa_composition_ecoli'. 
aa_composition_ecoli <- count(ecoli_prot_total, wordsize=1, alphabet=aa) 

# Show the composition of amino acids in the E.coli proteome
print(aa_composition_ecoli)

# Calculate, then show, the proportion of each amino acid in the E.coli proteome
aa_proportion_ecoli <- aa_composition_ecoli/sum(aa_composition_ecoli)
aa_proportion_ecoli <- round(aa_proportion_ecoli, digits = 3)
print(aa_proportion_ecoli)
```

```{r Determining the amino acid composition of A.aceti proteome}
# Follow the same code as outlined above for E.Coli
aaceti_prot_total <- unlist(aaceti_prot) 
summary(aaceti_prot_total)

aa_composition_aaceti <- count(aaceti_prot_total, wordsize=1, alphabet=aa)
print(aa_composition_aaceti)

aa_proportion_aaceti <- aa_composition_aaceti/sum(aa_composition_aaceti)
aa_proportion_aaceti <- round(aa_proportion_aaceti, digits = 3)
print(aa_proportion_aaceti)
```

```{r Installing and loading ggplot2 and ggpubr for creating graphs}
install.packages("ggplot2")
install.packages("ggpubr")
library("ggplot2")
library("ggpubr")
```

```{r Grouped bar chat showing nucleotide useage in E.coli and A.aceti}
#Create data frames from the nucleotide proportions, which are currently in table format, and add the column for Organism
dna_proportion_ecoli_df <- as.data.frame(dna_proportion_ecoli)
dna_proportion_ecoli_df$Organism <- c("E.coli")
dna_proportion_aaceti_df <- as.data.frame(dna_proportion_aaceti)
dna_proportion_aaceti_df$Organism <- c("A.aceti")

#Merge the two data frames together, to be one data frame
proportion_df <- rbind(dna_proportion_ecoli_df, dna_proportion_aaceti_df)
names(proportion_df) <- c("Nucleotide", "Proportion", "Organism") 
print(proportion_df)
```

Making the bar graphs
```{r Bar chart for nucleotide frequencies}
#Defining the variables for the grouped bar chart
Organism <- proportion_df$Organism
Nucleotide <- proportion_df$Nucleotide
Proportion <- proportion_df$Proportion

#Create the grouped bar chart
#The code specifies the dataframe to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A blue colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Blues".
ggbarplot(data=proportion_df, x="Nucleotide", y="Proportion", fill="Organism", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Blues") 
```


```{r Bar chat for amino acid frequencies}
#Create data frames from the aa proportions, which are currently in table format, and add the column for Organism
aa_proportion_ecoli_df <- as.data.frame(aa_proportion_ecoli)
aa_proportion_ecoli_df$Organism <- c("E.coli")
aa_proportion_aaceti_df <- as.data.frame(aa_proportion_aaceti)
aa_proportion_aaceti_df$Organism <- c("A.aceti")

#Merge the two data frames together, to be one data frame
aa_proportion_df <- rbind(aa_proportion_ecoli_df, aa_proportion_aaceti_df)
names(aa_proportion_df) <- c("Amino_Acids", "Proportion_aa", "Organism_aa") 
print(aa_proportion_df)
```

Making the bar graphs
```{r bar chart for amino acid frequencies}
#Defining the variables for the grouped bar chart
Organism_aa <- aa_proportion_df$Organism_aa
Amino_Acids <- aa_proportion_df$Amino_Acids
Proportion_aa <- aa_proportion_df$Proportion_aa

#Create the grouped bar chart
#The code specifies the dataframe to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A green colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Greens".
ggbarplot(data=aa_proportion_df, x="Amino_Acids", y="Proportion_aa", fill="Organism_aa", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Greens") 
```

# Describe any differences between the two organisms. 


# 5. Create a codon usage table and quantify the codon usage bias among all coding sequences. Describe any differences between the two organisms with respect to their codon usage bias. Provide charts to support your observations.

The RSCU stands for relative synonymous codon usage, which is a measure as to how preferred a codon of a particular amino acid is compared to the other codons which also encode it. A value of 1 indicating that the codon is used as expected. Values above 1 suggest that the codon is used more frequently than expected, while values below 1 indicate lower usage. For instance, codons like "aaa" and "att" show higher RSCU values, indicating they are used more often, whereas codons like "tga" and "tag" have lower values, suggesting less frequent use.

```{r Creating the codon usage table FIRST ATTEMPT, BUT THE ROWS ARE JUST ADDED TO EACH OTHER. SHOULD I HAVE INSTEAD ADDED JUST THE COLUMN DATA?}
# Create the codon usage data frame for E.coli, showing RSCU values. 
codons_ecoli_df <- uco(dna_ecoli,index="rscu",as.data.frame=TRUE)
print(codons_ecoli_df)

# Create the codon usage data frame for A.aceti, showing RSCU values
codons_aaceti_df <- uco(dna_aaceti,index="rscu",as.data.frame=TRUE)
print(codons_aaceti_df)
```

#### IF I HAVE TIME IT WOULD BE GOOD TO PUT THIS IN ONE TABLE, THOUGH IM NOT SURE IF THAT ONE TABLE WOULD BE EASY TO GRAPH TOGETHER OR NOT.
```{r ATTEMPT TO ADD THESE INTO ONE TABLE}
# Set the column names to include which organism it is

# OR CREATE A NEW DATAFRAME, WITH SELECTED FEATURES OF THE OTHER TWO, TO MAKE A DATAFRAME THAT HAS JUST THE AA, CODON, AND RSCU VALUE.

# Merge the rows required
```

```{r Grouped bar chart for RSCU values, minus 1}
# Add a column for organism name
codons_ecoli_df$Organism <- c("E.coli")

# Add a column for organism name
codons_aaceti_df$Organism <- c("A.aceti")

# Merge these dataframes into one
column_names <- colnames(codons_ecoli_df)
codons_df <- rbind(codons_ecoli_df, codons_aaceti_df)
names(codons_df) <- column_names
print(codons_df)

# Create a new column that calculates the RSCU value minus 1, such that values are either above or below a zero line.
codons_df$"RSCU-1" <- codons_df$RSCU-1

# Create a new column that contains the amino acid name plus it's codon
codons_df$"AA_and_codon" <- paste(codons_df$AA, ":", codons_df$codon)

# Put the data in alphabetical order, according to the amino acid 3 letter code, and print the resulting table
ordered_codons_df <- codons_df[order(codons_df$AA), ]
print(ordered_codons_df)
```

```{r Grouped barplot showing RSCU-1 values for E.coli and A.aceti}
# In the theme() component, I've set the column text size to 7, have adjusted the angle to 90 degrees,have justified the lables to be straight against the x-axis, and have shifted them left such that they line up under the tick lines. 
ggbarplot(data=ordered_codons_df, x="AA_and_codon", y="RSCU-1", fill="Organism", position=position_dodge (0.8)) +
  scale_fill_brewer(palette = "Oranges") +
  ggtitle("Relative Synonymous Codon Usage between E.coli and A.aceti") +
  xlab("Amino Acid : Codon") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust=1, vjust=0.5))
```




# 6. In the organism of interest, identify 10 protein sequence k-mers of length 3-5 which are the most over- and under-represented k-mers in your organism of interest. Are these k-mers also over- and under-represented in E. coli to a similar extent? Provide plots to support your observations. Why do you think these sequences are present at different levels in the genomes of these organisms?

```{r Extracting K-mers frequency for sizes 3, 4 and 5, and saving as separate data frames}
# Determine all K-mers of size 3, and their frequency of use
aaceti_kmer3 <- data.frame(count(aaceti_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer3) <- c("Kmer", "Frequency")

# Order this data, lowest to highest, then extract the top 10 and save in a new data frame called 'aaceti_kmer3_bottom10'
aaceti_kmer3 <- aaceti_kmer3[order(aaceti_kmer3$Frequency), ] 
aaceti_kmer3_bottom10 <- head(aaceti_kmer3, 10)
print(aaceti_kmer3_bottom10)

# Order this data, from highest use to lowest, then extract the top 10 and save in a new data frame called 'aaceti_kmer3_top10'
aaceti_kmer3 <- aaceti_kmer3[order(-aaceti_kmer3$Frequency), ] 
aaceti_kmer3_top10 <- head(aaceti_kmer3, 10)
print(aaceti_kmer3_top10)

# Repeat this process for all K-mers of size 4 amino acids
aaceti_kmer4 <- data.frame(count(aaceti_prot_total, wordsize=4, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer4) <- c("Kmer", "Frequency")

aaceti_kmer4 <- aaceti_kmer4[order(aaceti_kmer4$Frequency), ] # Orders by frequency, smallest to largest
aaceti_kmer4_bottom10 <- head(aaceti_kmer4, 10)
print(aaceti_kmer4_bottom10)

aaceti_kmer4 <- aaceti_kmer4[order(-aaceti_kmer4$Frequency), ] # Inclusion of - means this is now ordered by frequency, largest to smallest
aaceti_kmer4_top10 <- head(aaceti_kmer4, 10)
print(aaceti_kmer4_top10)

# Repeat this process for all K-mers of size 5 amino acids
aaceti_kmer5 <- data.frame(count(aaceti_prot_total, wordsize=5, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer5) <- c("Kmer", "Frequency")

aaceti_kmer5 <- aaceti_kmer5[order(aaceti_kmer5$Frequency), ] # Orders by frequency, smallest to largest
aaceti_kmer5_bottom10 <- head(aaceti_kmer5, 10)
print(aaceti_kmer5_bottom10)

aaceti_kmer5 <- aaceti_kmer5[order(-aaceti_kmer5$Frequency), ] # Inclusion of - means this is now ordered by frequency, largest to smallest
aaceti_kmer5_top10 <- head(aaceti_kmer5, 10)
print(aaceti_kmer5_top10)
```

Given that there are K-mers of size 4 and 5 not used in A.aceti, whereas K-mers of size 3 are all present in the A.aceti proteome, further comparisons between A.aceti and E.coli will focus just on the 10 most over- and under-represented k-mers of size 3.

```{r Extracting lists of the 10 most over- and under- represented 3 K-mers in A.aceti}
# Extract list of bottom 10 3-K-mers
Bottom10_kmer3 <- as.vector(aaceti_kmer3_bottom10$Kmer)
print(Bottom10_kmer3)

# Extract list of top 10 3-K-mers
Top10_kmer3 <- as.vector(aaceti_kmer3_top10$Kmer)
print(Top10_kmer3)
```

```{r Extracting frequencies of K-mers of size 3 in E.coli}
# Determine all K-mers of size 3, and their frequency of use, in E.coli
ecoli_kmer3 <- data.frame(count(ecoli_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(ecoli_kmer3) <- c("Kmer", "Frequency")
```

```{r}
# Subset the E.coli list, using the top 10 A.aceti list, and save as a new df
ecoli_kmer3_topsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Top10_kmer3, ]

# Add a column for Organism, and note is as E.coli
ecoli_kmer3_topsubset$Organism <- c("E.coli")
print(ecoli_kmer3_topsubset)

# Subset the E.coli list, using the bottom 10 A.aceti list, and save as a new df
ecoli_kmer3_bottomsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Bottom10_kmer3, ]

# Add a column for Organism, and note is as E.coli
ecoli_kmer3_bottomsubset$Organism <- c("E.coli")
print(ecoli_kmer3_bottomsubset)
```

```{r}
#Prepare the A.aceti K-mer 3 data frames, by adding a column for organism
aaceti_kmer3_bottom10$Organism <- c("A.aceti")
aaceti_kmer3_top10$Organism <- c("A.aceti")

# Merge the two organisms' data frames that contain the frequency data for the top 10 3-K-mers from A.aceti
column_names_kmer3 <- colnames(aaceti_kmer3_bottom10)

Top_kmer3_df <- rbind(aaceti_kmer3_top10, ecoli_kmer3_topsubset)
names(Top_kmer3_df) <- column_names_kmer3
print(Top_kmer3_df)

# Merge the two organisms' data frames that contain the frequency data for the bottom 10 3-K-mers from A.aceti
Bottom_kmer3_df <- rbind(aaceti_kmer3_bottom10, ecoli_kmer3_bottomsubset)
names(Bottom_kmer3_df) <- column_names_kmer3
print(Bottom_kmer3_df)
```

```{r Boxplot comparison of A.aceti top 10 k-mers}
ggbarplot(data=Top_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Greens") 
```

```{r Boxplot comparison of A.aceti bottom 10 k-mers}
ggbarplot(data=Bottom_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Reds") 
```



```{r Recording Session Info}
sessionInfo()
```