---
title: "SLE777 Bioinformatics Assessment 4"
author: "Renee Carroll, s225723235"
date: "2025-10-5"
output: 
  html_document: 
    toc: true
    fig_width: 7
    fig_height: 7
theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("car")
#install.packages("seqinr")
#install.packages("R.utils")
#install.packages("ggplot2")
#install.packages("ggpubr")
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(seqinr))
suppressPackageStartupMessages(library(R.utils))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
```

# Part 1

## Analysis of RNA-seq Expression Data


#### 1.1. Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes.


```{r download files from GitHub, and show first 6 genes, echo=FALSE}
URL1 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL1, destfile="gene_expression.tsv")
gene_expression <- read.delim("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t", stringsAsFactors = FALSE)
```


The gene expression data for the first six genes for each of the three samples of interest is shown in the table below.


```{r first 6 genes, echo=FALSE}
knitr::kable(head(gene_expression),
             caption = "Table 1. Gene expression levles for the first six genes")
```


#### 1.2. Make a new column which is the mean of the other columns. Show a table of values for the first six genes.


The new column is added to the end of the data frame, and the updated table is shown below.


```{r mean saved as new column, echo=FALSE}
gene_expression$mean_expression <- rowMeans(gene_expression)
knitr::kable(head(gene_expression),
             caption = "Table 2. Gene expression levles for the first six genes, including mean values")
```


#### 1.3. List the 10 genes with the highest mean expression


```{r 10 genes with highest mean, echo=FALSE}
ordered_expression <- gene_expression[order(-gene_expression$mean_expression),]
genes_highest_expression <- head(ordered_expression, 10)
```


The 10 genes with the highest mean expression are:


1. `r rownames(genes_highest_expression)[1]`

2. `r rownames(genes_highest_expression)[2]`

3. `r rownames(genes_highest_expression)[3]`

4. `r rownames(genes_highest_expression)[4]`

5. `r rownames(genes_highest_expression)[5]`

6. `r rownames(genes_highest_expression)[6]`

7. `r rownames(genes_highest_expression)[7]`

8. `r rownames(genes_highest_expression)[8]`

9. `r rownames(genes_highest_expression)[9]`

10. `r rownames(genes_highest_expression)[10]`


#### 1.4. Determine the number of genes with a mean < 10


```{r number of genes with low expression < 10, echo=FALSE}
low_exp_genes <- subset(gene_expression, mean_expression<10)
```


The number of genes with a mean expression <10 is `r nrow(low_exp_genes)`.


#### 1.5. Make a histogram plot of the mean values and include it into your report.


Histograms area useful way to visualize expression data. Figure 1 below shows the histogram of mean global gene expression from the three samples of interest.

```{r histogram of expression values, fig.cap="*Figure 1. Mean global gene expression from three samples of interest, showing severe right skewing in the data, indicating a large proportion of genes with low expression levels.*", echo=FALSE}
gene_exp_greaterthan0 <- subset(gene_expression, mean_expression>0)
hist(gene_exp_greaterthan0$mean_expression, breaks=100, main = "Mean Gene Expression", xlab = "Mean Gene Expression", col = "dodgerblue2")
```


From this histogram, we can see that the vast majority of genes have very low expression. This severe right skewing of the data makes it difficult to glean useful information from this graph, as the large number of lowly expression genes dominates.

Data with a distribution such as this is better visualized on a logarithmic scale.

As there would be genes in this data with mean expression levels below 1, we can add 1 to all mean expression values, to ensure that the log of the values stays positive.

Here is the resulting histogram.


```{r histogram of log mean values, fig.cap="*Figure 2. Mean global gene expression from three samples of interest, on a logarithmic scale.*", echo=FALSE}
log_gene_exp_greaterthan0 <- log(gene_exp_greaterthan0+1)
hist(log_gene_exp_greaterthan0$mean_expression, breaks=50, main = "Mean Gene Expression (log)", xlab = "Log Mean Gene Expression", col = "dodgerblue2")
```


# Part 1
## Tree Circumference Data Analysis


#### 1.6. Import this csv file into an R object. 


The file “growth_data.csv” contains measurements for the circumferences of trees growing at two sites, a control site and a treatment site, which were planted 20 years ago.

```{r Import csv file, echo=FALSE}
URL2 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
download.file(URL2, destfile="growth_data.csv")
growth_data <- read.csv("growth_data.csv") # read.csv automatically treats 
```


#### What are the column names?


The column names of this data set are:

1. `r colnames(growth_data)[1]`

2. `r colnames(growth_data)[2]`

3. `r colnames(growth_data)[3]`

4. `r colnames(growth_data)[4]`

5. `r colnames(growth_data)[5]`

6. `r colnames(growth_data)[6]`


#### 1.7. Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.


```{r Create new data frame for the mean and sd, echo=FALSE}
girth_stats_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  Mean_2005_cm = c(mean(subset(growth_data, Site == "northeast")$Circumf_2005_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  SD_2005_cm = c(sd(subset(growth_data, Site == "northeast")$Circumf_2005_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  Mean_2020_cm = c(mean(subset(growth_data, Site == "northeast")$Circumf_2020_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2020_cm)),
  SD_2020_cm = c(sd(subset(growth_data, Site == "northeast")$Circumf_2020_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2020_cm))
)
girth_stats_df[,2:5] <- round(girth_stats_df[,2:5], digits = 2)
```


The following table shows the mean and standard deviation of tree circumference at the start and end of the study at both sites. For ease of data viewing, all values contained in columns 2 to 5 have been rounded to 2 decimal places.

``` {r Print the new data frame of mean and sd, echo=FALSE}
knitr::kable(girth_stats_df,
             caption = "Table 3. Girth statistics at the start and end of study")
```


#### 1.8. Make a box plot of tree circumference at the start and end of the study at both sites.


```{r boxplot of tree circumference at the start and end of the study at both sites, fig.cap="*Figure 3. Boxplot of tree girth measurements (in cm) taken in 2005 and 2020 at the Norheast and Southwest study sites. The median value is shown by the black horizontal central line. The interquartile range, comprised of the upper quartile above the median, and the lower quartile below the median, is represented by the coloured box. The maximum and minimum values are shown by the ends of the whiskers.*", echo=FALSE}
boxplot(
  subset(growth_data, Site == "northeast")$Circumf_2005_cm, 
  subset(growth_data, Site == "northeast")$Circumf_2020_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2005_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2020_cm, 
  names = c("Northeast 2005", "Northeast 2020", "Southwest 2005", "Southwest 2020"), 
  col = c("forestgreen"),
  ylab = "Circumference (cm)", 
  main = "Increase in tree girth between 2005 and 2020, at each study site")
```


#### 1.9. Calculate the mean growth over the last 10 years at each site.


```{r mean growth over the last 10 years, echo=FALSE}
# Add a column to the existing data frame, to contain the calculation of 2020 measurements minus the 2010 measurements.
growth_data$growth_2010_2020 <- growth_data$Circumf_2020_cm-growth_data$Circumf_2010_cm 

# Create a new data frame to present the mean of the 10 year growth data, for each location.
mean_10yr_growth_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  "Mean 10 year growth (cm)" = c(mean(subset(growth_data, Site == "northeast")$growth_2010_2020), mean(subset(growth_data, Site == "southwest")$growth_2010_2020)),
  check.names = FALSE 
)
```


Table 4 summarises the mean growth, measured in cm, over the last 10 years at each of the sites.


```{r table showing mean growth over 10 years, echo=FALSE}
knitr::kable(mean_10yr_growth_df,
             caption = "Table 4. Mean growth over last 10 years")
```


#### 1.10. Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.

Prior to conducting a t-test, it is important to fist check for equal variance in the data, and one way to achieve this is to conduct the Levene's Test.


``` {r, echo=FALSE}
leveneTest(growth_2010_2020 ~ Site, data = growth_data)
```


As the Levene's Test returned a P-value of 0.006116, it indicates the variances in this data are not equal, and therefore the appropriate t-test to use is a Welch's t-test, as it adjusts for unequal variances.


```{r t.test, echo=FALSE}
t.test(growth_2010_2020 ~ Site, data = growth_data, var.equal = FALSE) # var.equal = FALSE isn't required here, as it is the default for t.test() in R.
```


This t-test provided a p-value of 0.06229, indicating the difference in 10 year growth between the two sites is not statistically significant, however as this value is only slightly above the cutoff of 0.05, is is close to being significant.

The recommendation is to keep monitoring tree growth at the two sites, and repeat the analysis in the future.


# Part 2
## Examining biological sequence diversity between E.coli and A.aceti


Here in this report, we will perform a comparative analysis on the coding sequences of two bacteria, Escherichia coli (E.coli) and Acetobacter aceti (A.aceti).


#### 2.1.1. Download the whole set of coding DNA sequences for E. coli and your organism of interest. How many coding sequences are present in these organisms? Present this in the form of a table. 


```{r Download E.coli coding DNA sequences, echo=FALSE}
# Download the E.coli cds, and unzip the file.
URL3 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_0_collection/escherichia_coli_str_k_12_substr_mg1655_gca_000005845/cds/Escherichia_coli_str_k_12_substr_mg1655_gca_000005845.ASM584v2.cds.all.fa.gz"
#download.file(URL3, destfile="ecoli_cds.fa.gz")
#gunzip("ecoli_cds.fa.gz")

# Download the cds for my organism of interest, Acetobacter aceti GCA_002005445, and unzip the file.
URL4 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_80_collection/acetobacter_aceti_gca_002005445/cds/Acetobacter_aceti_gca_002005445.ASM200544v1.cds.all.fa.gz"
#download.file(URL4, destfile="aaceti_cds.fa.gz")
#gunzip("aaceti_cds.fa.gz")

# Read in the two sequence files.
ecoli_cds <- seqinr::read.fasta("ecoli_cds.fa")
aaceti_cds <- seqinr::read.fasta("aaceti_cds.fa")
```

```{r Table of the number of coding sequences present in these organisms, echo=FALSE}
cds_df <- data.frame(
  Organism = c("E.coli", "A.aceti"),
  "Number of Coding Sequences" = c(length(ecoli_cds), length(aaceti_cds)),
  check.names = FALSE 
  )
knitr::kable(cds_df,
             caption = "Table 5. Number of coding sequences present in E.coli and A.aceti")
```


#### 2.1.2. Describe any differences between the two organisms. 


The organism E.coli has `r (length(ecoli_cds)-length(aaceti_cds))` additional genes in its coding DNA, which is an increase of `r round((length(ecoli_cds)-length(aaceti_cds))/length(aaceti_cds)*100, 1)`%. 

This data suggests that E.coli is a more complex organism than A.aceti, as a greater number of genes would result in a greater number of proteins, and hence additional or more complex biological functionality.


#### 2.2.1 How much coding DNA is there in total for these two organisms? Present this in the form of a table.


```{r Table of the total amount of coding DNA for the two organisms, echo=FALSE}
length_ecoli_genes <- as.numeric(summary(ecoli_cds)[,1])
length_aaceti_genes <- as.numeric(summary(aaceti_cds)[,1])
cds_total_df <- data.frame(
  Organism = c("E.coli", "A.aceti"), 
  "Total Amount of Coding Sequence" = c(sum(length_ecoli_genes), sum(length_aaceti_genes)),
  check.names = FALSE 
) # Including the check.names = FALSE code so that R doesn't add '.' in between the words of the Column names.
knitr::kable(cds_total_df,
             caption = "Table 6. Total quantity of coding DNA in E.coli and A.aceti")
```


#### 2.2.2. Describe any differences between the two organisms. 


In line with the previous observation that E.coli contains `r round((length(ecoli_cds)-length(aaceti_cds))/length(aaceti_cds)*100, 1)`% more genes than A.aceti, we can further determine that E.coli has a proportinal increase in the total amount of coding sequence as well, namely, that E.coli contains `r  round((sum(length_ecoli_genes)-sum(length_aaceti_genes))/sum(length_aaceti_genes)*100, 1)`% more coding DNA in total. This further supports the hypothesis that E.coli is a more complex organism than A.aceti.


#### 2.3.1. Calculate the length of all coding sequences in these two organisms. Make a boxplot of coding sequence length in these organisms. 


```{r Boxplot of coding sequence length in these organisms, fig.cap="*Figure 4. Boxplot of coding sequence lengths between two organisms, E.coli (red) and A.aceti (blue). The median values are shown by the black horizontal central lines. The interquartile range, comprised of the upper quartile above the median, and the lower quartile below the median, are represented by the coloured boxes. The maximum and minimum values are shown by the ends of the whiskers, with outlying values shown by points above the whisker border line.*", echo=FALSE}
boxplot(length_ecoli_genes, length_aaceti_genes, names = c("E.coli", "A.aceti"), main = "Coding DNA Sequence Lengths in E.coli and A.aceti", ylab="Length of cDNA (bp)", col=c("red", "blue"))
```


#### 2.3.2 What is the mean and median coding sequence length of these two organisms? Describe any differences between the two organisms. 


The average length of a coding sequence in E.coli is `r round(mean(length_ecoli_genes), 0)` bp long, and in A.aceti is `r round(mean(length_aaceti_genes), 0)` bp long. The median coding sequence length in E.coli is `r median(length_ecoli_genes)`, and in A.aceti is `r median(length_aaceti_genes)`. 

Together with the boxplot shows in Figure 4 above, this data reveals that the transcriptomes of E.coli and A.aceti are comparable in terms of gene length. In both organisms, the vast majority of coding sequences are around the 800-900 bp length, with both organisms having a smaller number of very large genes, in the approximate range of 2500-8000 bp long, as shown by the data points above the plots' upper whisker lines. The presence of these large genes are the reason the calculated average length is larger than the calculated median length, in both organisms. 


#### 2.4.1. Calculate the frequency of DNA bases in the total coding sequences for both organisms. Create a bar plot for nucleotide frequency. 


```{r Nucleotide composition and frequencies in E.coli and A.aceti, echo=FALSE}

# Determining DNA composition for each organism
dna_ecoli <- unlist(ecoli_cds) 
dna_composition_ecoli <- count(dna_ecoli,1)

dna_aaceti <- unlist(aaceti_cds) 
dna_composition_aaceti <- count(dna_aaceti,1) 

# Determining DNA proportion for each organism
dna_proportion_ecoli <- dna_composition_ecoli/sum(dna_composition_ecoli)
dna_proportion_ecoli <- round(dna_proportion_ecoli, digits = 3) # digits=3 means that the result is rounded to 3 decimal places.

dna_proportion_aaceti <- dna_composition_aaceti/sum(dna_composition_aaceti)
dna_proportion_aaceti <- round(dna_proportion_aaceti, digits = 3)
```

``` {r Creating the bar chart for nucleotide frequencies using ggplot2, fig.cap="*Figure 5. Grouped bar chat showing the frequencies of nucleotide composition in the two organisms, E.coli and A.aceti.*", echo=FALSE}
# Create data frames from the nucleotide proportions, and add the column for Organism
dna_proportion_ecoli_df <- as.data.frame(dna_proportion_ecoli)
dna_proportion_ecoli_df$Organism <- c("E.coli")

dna_proportion_aaceti_df <- as.data.frame(dna_proportion_aaceti)
dna_proportion_aaceti_df$Organism <- c("A.aceti")

# Merge the two data frames together, to create one data frame for the bar plot
proportion_df <- rbind(dna_proportion_ecoli_df, dna_proportion_aaceti_df)
names(proportion_df) <- c("Nucleotide", "Proportion", "Organism") 

# Define the variables for the grouped bar chart
Organism <- proportion_df$Organism
Nucleotide <- proportion_df$Nucleotide
Proportion <- proportion_df$Proportion

# Create the grouped bar chart. 
# The code specifies the data frame to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A blue colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Blues".
ggbarplot(data=proportion_df, x="Nucleotide", y="Proportion", fill="Organism", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Blues") 
```


#### 2.4.2. Describe any differences between the two organisms.


While nucleotide composition is comparable between E.coli and A.aceti, it is observed that the coding genome of the organism of interest, A.aceti, contains an decreased proportion of adenine (a) and thymine (t), and an increased proportion of cytosine (c) and guanine (g), as compared to the coding genome of E.coli. 


#### 2.4.3. Calculate the frequency of amino acids in the total protein sequences for both organisms. Create a bar plot for amino acid frequency. 


```{r Determining the frequency of amino acids in the total protein sequences of each organism, echo=FALSE}
# Translate the cDNA for each organism
ecoli_prot <- lapply(ecoli_cds, translate)
aaceti_prot <- lapply(aaceti_cds, translate)

# Unlist the protein sequences and saving then as a single vector
ecoli_prot_total <- unlist(ecoli_prot)
aaceti_prot_total <- unlist(aaceti_prot) 

# Creates the list of possible amino acids to form our alphabet to use in the 'count' code.
aa <- unique(ecoli_prot_total) 

# Removes the stop codon asterisks.
aa <- aa[aa != "*"] 

# Count the number of each amino acid in each organisms proteome vector, in combinations of 1's (ie. single amino acids), using the aa alphabet previously determined, and save it as a new object. 
aa_composition_ecoli <- count(ecoli_prot_total, wordsize=1, alphabet=aa) 
aa_composition_aaceti <- count(aaceti_prot_total, wordsize=1, alphabet=aa)

# Calculate the proportion of each amino acid in each organisms' proteome
aa_proportion_ecoli <- aa_composition_ecoli/sum(aa_composition_ecoli)
aa_proportion_ecoli <- round(aa_proportion_ecoli, digits = 3)
aa_proportion_aaceti <- aa_composition_aaceti/sum(aa_composition_aaceti)
aa_proportion_aaceti <- round(aa_proportion_aaceti, digits = 3)
```

```{r Creating the bar chart for amino acid frequencies using ggplot2, fig.cap="*Figure 5. Grouped bar chat showing the frequencies of amino acid composition in the two organisms, E.coli and A.aceti.*", echo=FALSE}
#Create data frames from the aa proportions, and add the column for Organism
aa_proportion_ecoli_df <- as.data.frame(aa_proportion_ecoli)
aa_proportion_ecoli_df$Organism <- c("E.coli")

aa_proportion_aaceti_df <- as.data.frame(aa_proportion_aaceti)
aa_proportion_aaceti_df$Organism <- c("A.aceti")

#Merge the two data frames together, to create one data frame for the bar plot
aa_proportion_df <- rbind(aa_proportion_ecoli_df, aa_proportion_aaceti_df)
names(aa_proportion_df) <- c("Amino_Acids", "Proportion_aa", "Organism_aa") 

#Defining the variables for the grouped bar chart
Organism_aa <- aa_proportion_df$Organism_aa
Amino_Acids <- aa_proportion_df$Amino_Acids
Proportion_aa <- aa_proportion_df$Proportion_aa

#Create the grouped bar chart
#The code specifies the data frame to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A green colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Greens").
ggbarplot(data=aa_proportion_df, x="Amino_Acids", y="Proportion_aa", fill="Organism_aa", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Greens") 
```


#### 2.4.4. Describe any differences between the two organisms. 


The grouped bar chart in Figure 5 reveals that the proportion of each amino acid comprising the proteome of each organism, E.coli and A.aceti, is very comparable. There are only marginal differences, most notably observed with the slightly increased use of A (Alanine), G (Glycine), P (Proline), and R (Arginine) in A.aceti versus E.coli, and a decreased use of I (Isoleucine), K (Lysine), N (Asparagine), and Q (Glutamine).  


#### 2.5.1. Create a codon usage table and quantify the codon usage bias among all coding sequences. 


Given that there is degeneracy in the DNA code, in that the majority of amino acids can be coded for by more than one codon, it is interesting to investigate whether organisms show any preferential usage for particular codons for the same amino acid, and to compare this between species. 

Table 7 below shows the total codon counts (under heading 'eff'), the relative frequencies (under heading 'freq'), and the calculated RSCU values (under heading 'RSCU'). 

RSCU stands for Relative Synonymous Codon Usage, and it is a measure that tells us whether there is any preferential usage of a particular codon for a particular amino acid, as compared to the relative usage of other codons for that same amino acid. 


```{r Creating the codon usage table, echo=FALSE}
# Create the codon usage data frame for E.coli, showing RSCU values. 
codons_ecoli_df <- uco(dna_ecoli, index = "rscu", as.data.frame=TRUE)
codons_ecoli_table <- codons_ecoli_df
colnames(codons_ecoli_table) <- c("AA", "codon", "E.coli_eff", "E.coli_freq", "E.coli_RSCU")

# Create the codon usage data frame for A.aceti, showing RSCU values
codons_aaceti_df <- uco(dna_aaceti, index = "rscu", as.data.frame=TRUE)
codons_aaceti_table <- codons_aaceti_df
colnames(codons_aaceti_table) <- c("AA", "codon", "A.aceti_eff", "A.aceti_freq", "A.aceti_RSCU")

# Creating a new data frame from 'codons_ecoli_df' data frame, and adding the A.aceti columns from the 'codons_aaceti_df' data frame.
codons_table <- codons_ecoli_table
codons_table$A.aceti_eff <- codons_aaceti_table$A.aceti_eff
codons_table$A.aceti_freq <- codons_aaceti_table$A.aceti_freq
codons_table$A.aceti_RSCU <- codons_aaceti_table$A.aceti_RSCU
knitr::kable(codons_table,
             row.names = FALSE,
             caption = "Table 7. Codon usage statistics of E.coli and A.aceti. AA: Amino Acid three leter code, codon: encoding codon, eff: total codon counts, freq: relative frequencies, RSCU: Relative Synonymous Codon Usage.")
```


#### 2.5.2. Provide charts to support your observations.


The RSCU value is useful to focus on when comparing codon preference between organisms. 
An RSCU value of 1 means the codon is used as expected, values above 1 indicate the codon is being used more frequently than expected, and values below 1 indicate the codon is being used less frequently. 

The RSCU values can be better visualized in the bar plots in Figures 6(a) and 6(b) below, where the data has been split approximately evenly, for enhanced viewing.


```{r Creating the data frame which will be used to create the grouped bar chart for RSCU values, echo=FALSE}
# Using the original codon data frames, add a column for organism name
codons_ecoli_df$Organism <- c("E.coli")
codons_aaceti_df$Organism <- c("A.aceti")

# Merge these data frames into one
column_names <- colnames(codons_ecoli_df) # creates the vector of column names
codons_df <- rbind(codons_ecoli_df, codons_aaceti_df) # combines the rows of data
names(codons_df) <- column_names # assigns the column names of the new data frame to be the names from the set vector

# Create a new column that calculates the RSCU value minus 1, such that values are either above or below a zero line.
codons_df$"RSCU-1" <- codons_df$RSCU-1

# Create a new column that contains the amino acid name plus it's codon
codons_df$"AA_and_codon" <- paste(codons_df$AA, ":", codons_df$codon)

# Put the data in alphabetical order, according to the amino acid 3 letter code
ordered_codons_df <- codons_df[order(codons_df$AA), ]
```

```{r Grouped barplot (a) showing RSCU-1 values, fig.cap="*Figure 6(a). Grouped bar chat showing Relative Synonymous Codon Usage in E.coli and A.aceti, for the first subset of amino acids. RSCU values minus 1 are shown on the y axis, and on the x axis is each of the amino acids and their various respective codons.*", echo=FALSE}
# Subsetting the data frame using [] after the data frame name, to select rows 1 to 70 to graph.
# In the theme() component, I've set the column text size to 10, have adjusted the angle to 90 degrees, have justified the labels to be straight against the x-axis, and have shifted them left such that they line up under the tick lines. 
ggbarplot(data=ordered_codons_df[1:70, ], x="AA_and_codon", y="RSCU-1", fill="Organism", position=position_dodge (0.8)) +
  scale_fill_brewer(palette = "Oranges") +
  ggtitle("Relative Synonymous Codon Usage between E.coli and A.aceti") +
  xlab("Amino Acid : Codon") +
  theme(axis.text.x = element_text(size = 10, angle = 90, hjust=1, vjust=0.5))
```

```{r Grouped barplot (b) showing RSCU-1 values, fig.cap="*Figure 6 (b). Grouped bar chat showing Relative Synonymous Codon Usage in E.coli and A.aceti, for the second subset of amino acids. RSCU values minus 1 are shown on the y axis, and on the x axis is each of the amino acids and their various respective codons. Note that 'Stp' refers to stop codons, and not a specific amino acid.*", echo=FALSE}
# Subsetting the data frame using [] after the data frame name, to select rows 71 to 128 to graph.
# In the theme() component, I've set the column text size to 10, have adjusted the angle to 90 degrees, have justified the labels to be straight against the x-axis, and have shifted them left such that they line up under the tick lines. 
ggbarplot(data=ordered_codons_df[71:128, ], x="AA_and_codon", y="RSCU-1", fill="Organism", position=position_dodge (0.8)) +
  scale_fill_brewer(palette = "Oranges") +
  ggtitle("Relative Synonymous Codon Usage between E.coli and A.aceti") +
  xlab("Amino Acid : Codon") +
  theme(axis.text.x = element_text(size = 10, angle = 90, hjust=1, vjust=0.5))
```


#### 2.5.3. Describe any differences between the two organisms with respect to their codon usage bias. 


Figure 6 (a) and (b) reveals that while there are many codons being used to similar frequencies in E.coli and A.aceti, such as the Leucine codon 'ctg', there are also some interesting differences, such as the preferred codon usage for Serine, which in A.aceti is 'tcc' and in E.coli is 'agc'. Also interesting is the preferred different in Stop codon usgae, as denoted by 'Stp' in Figure 6 (b), with A.aceti preferencing the codon 'tga' and E.coli preferencing the codon 'taa'.

Such differences in codon preference provide interesting insights into the evolutionary histories of these two bacteria.


#### 2.6.1. In the organism of interest, identify 10 protein sequence k-mers of length 3-5 which are the most over- and under-represented k-mers in your organism of interest. Are these k-mers also over- and under-represented in E. coli to a similar extent? 


K-mers are short substrings of biological sequence of a fixed length, denoted by 'k'. 
Here, we'll investigate the presence of protein sequence k-mers in the total protein sequences of E.coli and A.aceti, of length 3, 4 and 5 amino acids.

These calculations can be seen in in Tables 8 to 13 below.


```{r Extracting K-mers frequency for sizes 3, 4 and 5, and saving as separate data frames, echo=FALSE}
# Determine all K-mers of size 3, and their frequency of use
aaceti_kmer3 <- data.frame(count(aaceti_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer3) <- c("Kmer", "Frequency")

# Order this data by frequency, highest to lowest, then extract the top 10 and bottom 10 rows, to save in new data frames
aaceti_kmer3 <- aaceti_kmer3[order(-aaceti_kmer3$Frequency), ] # the minus sign before the data frame name makes the ordering highest to lowest.
aaceti_kmer3_top10 <- head(aaceti_kmer3, 10)
aaceti_kmer3_bottom10 <- tail(aaceti_kmer3, 10)

# Repeat this process for all K-mers of size 4 amino acids
aaceti_kmer4 <- data.frame(count(aaceti_prot_total, wordsize=4, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer4) <- c("Kmer", "Frequency")

# Order this data by frequency, highest to lowest, then extract the top 10 and bottom 10 rows, to save in new data frames
aaceti_kmer4 <- aaceti_kmer4[order(-aaceti_kmer4$Frequency), ] # the minus sign before the data frame name makes the ordering highest to lowest.
aaceti_kmer4_top10 <- head(aaceti_kmer4, 10)
aaceti_kmer4_bottom10 <- tail(aaceti_kmer4, 10)

# Repeat this process for all K-mers of size 5 amino acids
aaceti_kmer5 <- data.frame(count(aaceti_prot_total, wordsize=5, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer5) <- c("Kmer", "Frequency")

# Order this data by frequency, highest to lowest, then extract the top 10 and bottom 10 rows, to save in new data frames
aaceti_kmer5 <- aaceti_kmer5[order(-aaceti_kmer5$Frequency), ] # the minus sign before the data frame name makes the ordering highest to lowest.
aaceti_kmer5_top10 <- head(aaceti_kmer5, 10)
aaceti_kmer5_bottom10 <- tail(aaceti_kmer5, 10)

# Create all tables
knitr::kable(aaceti_kmer3_top10,
             row.names = FALSE,
             caption = "Table 8. Ten most frequently used amino acid 3-mers in A.aceti.")

knitr::kable(aaceti_kmer3_bottom10,
             row.names = FALSE,
             caption = "Table 9. Ten least frequently used amino acid 3-mers in A.aceti.")

knitr::kable(aaceti_kmer4_top10,
             row.names = FALSE,
             caption = "Table 10. Ten most frequently used amino acid 4-mers in A.aceti.")

knitr::kable(aaceti_kmer4_bottom10,
             row.names = FALSE,
             caption = "Table 11. Ten least frequently used amino acid 4-mers in A.aceti.")

knitr::kable(aaceti_kmer5_top10,
             row.names = FALSE,
             caption = "Table 12. Ten most frequently used amino acid 5-mers in A.aceti.")

knitr::kable(aaceti_kmer5_bottom10,
             row.names = FALSE,
             caption = "Table 13. Ten least frequently used amino acid 5-mers in A.aceti.")
```


This analysis reveals that there are many 4-mers and 5-mers which are not present at all in the proteome of A.aceti, as revealed by frequencies of zero in Tables 11 and 13 above. 

In addition, the frequency values returned for the top most used 3-mers are all larger than the frequency values for the top most used 4-mers and 5-mers. This is as expected, given that increased length of k-mer also brings increased specialty, and and hence is present at a reduced frequency within the same data set.

For these reasons, further comparisons between A.aceti and E.coli will focus just on 3-mers.


```{r Extracting lists of the 10 most over- and under- represented 3 K-mers in A.aceti, echo=FALSE}
Top10_kmer3 <- as.vector(aaceti_kmer3_top10$Kmer)
Bottom10_kmer3 <- as.vector(aaceti_kmer3_bottom10$Kmer)
```

```{r Extracting frequency data of 3-mers in E.coli proteome, echo=FALSE}
ecoli_kmer3 <- data.frame(count(ecoli_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(ecoli_kmer3) <- c("Kmer", "Frequency")
```

```{r Subsetting the E.coli lists of 3-mers, echo=FALSE}
# Subset the E.coli list, using the top 10 A.aceti list, and save as a new df
# This code tells R to save to a new data frame all columns of data from rows in the 'ecoli_kmer3' data frame, where the values in the 'Kmer' column match a value in the 'Top10_kmer3' vector. 
ecoli_kmer3_topsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Top10_kmer3, ] 
ecoli_kmer3_topsubset$Organism <- c("E.coli") # Add a column for Organism, and note it as E.coli

# Subset the E.coli list, using the bottom 10 A.aceti list, and save as a new df
ecoli_kmer3_bottomsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Bottom10_kmer3, ]
ecoli_kmer3_bottomsubset$Organism <- c("E.coli") # Add a column for Organism, and note it as E.coli
```


#### 2.6.2. Provide plots to support your observations of k-mer usage between E.coli and your organism of interest. 

See Figures 7 and 8 below for the grouped bar plots showing top-10 and bottom 10 3-mers analysis.


```{r Preparing the data frame, echo=FALSE}
#Prepare the A.aceti K-mer 3 data frames, by adding a column for organism
aaceti_kmer3_bottom10$Organism <- c("A.aceti")
aaceti_kmer3_top10$Organism <- c("A.aceti")

# Merge the two organisms' data frames that contain the frequency data for the top 10 3-K-mers from A.aceti
column_names_kmer3 <- colnames(aaceti_kmer3_bottom10) # Sets what the column names will be
Top_kmer3_df <- rbind(aaceti_kmer3_top10, ecoli_kmer3_topsubset)
names(Top_kmer3_df) <- column_names_kmer3

# Merge the two organisms' data frames that contain the frequency data for the bottom 10 3-K-mers from A.aceti
Bottom_kmer3_df <- rbind(aaceti_kmer3_bottom10, ecoli_kmer3_bottomsubset)
names(Bottom_kmer3_df) <- column_names_kmer3
```

```{r Boxplot comparison of A.aceti top 10 k-mers, fig.cap="*Figure 7. Grouped bar chat showing frequency of top-10 3-mers in A.aceti, and their relative occurence in the E.coli proteome.*", echo=FALSE}
# This line keeps the levels in the same order as they appear in the table, and not graphed in alphabetical order, as is default for ggplot. As there are duplicates in the Kmer column, the 'levels = ' code specifies for only unique factors.
Top_kmer3_df$Kmer <- factor(Top_kmer3_df$Kmer, levels = unique(Top_kmer3_df$Kmer))
ggbarplot(data=Top_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  ggtitle("Ten most over-represented amino acid 3-mers in A.aceti, \n with comparison to frequency of presentation in E.coli") +
  scale_fill_brewer(palette = "Greens") 
```

```{r Boxplot comparison of A.aceti bottom 10 k-mers, fig.cap="*Figure 8. Grouped bar chat showing frequency of bottom-10 3-mers in A.aceti, and their relative occurence in the E.coli proteome.*", echo=FALSE}
# This line keeps the levels in the same order as they appear in the table, and not graphed in alphabetical order, as is default for ggplot. As there are duplicates in the Kmer column, the 'levels = ' code specifies for only unique factors.
Bottom_kmer3_df$Kmer <- factor(Bottom_kmer3_df$Kmer, levels = unique(Bottom_kmer3_df$Kmer))
ggbarplot(data=Bottom_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  ggtitle("Ten most under-represented amino acid 3-mers in A.aceti, \n with comparison to frequency of presentation in E.coli") +
  scale_fill_brewer(palette = "Reds") 
```


#### 2.6.3. Why do you think these sequences are present at different levels in the genomes of these organisms?


Figures 7 and 8 show the ten most over- and under-represented 3-mers in A.aceti, respectively, with their corresponding frequencies observed in E.coli as comparison. These bar plots clearly show differences in observed frequencies between the two organisms, with E.coli having lower representation of 7/10 of A.aceti's most over-represented 3-mers, and higher representation of 9/10 of A.aceti's most under-represented 3-mers.

However, in order to clearly visualize the relative difference of these differences in frequencies, it's helpful to see the data on the same bar plot, as shown below in Figure 9.

```{r Boxplot comparison of A.aceti bottom and top 10 k-mers, fig.cap="*Figure 9. Grouped bar chat showing frequency of bottom-10 3-mers and top-10 kmers in A.aceti, and their relative occurence in the E.coli proteome.*", echo=FALSE}
# Combine the two data frames, and order first by organism, then by frequency
combined_kmer3_df <- rbind(Top_kmer3_df, Bottom_kmer3_df)
combined_kmer3_df <- combined_kmer3_df[order(combined_kmer3_df$Organism, combined_kmer3_df$Frequency), ]

# This first line again keeps the order as per the data frame, and not alphabetical.
combined_kmer3_df$Kmer <- factor(combined_kmer3_df$Kmer, levels = unique(combined_kmer3_df$Kmer))
ggbarplot(data=combined_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  ggtitle("Ten most under- and over-represented amino acid 3-mers in A.aceti, \n with comparison to frequency of presentation in E.coli") +
  scale_fill_brewer(palette = "Purples")+
  theme(axis.text.x = element_text(size = 10, angle = 90, hjust=1, vjust=0.5))
```


While there are observable differences to 3-mer representation, as shown in Figures 7 and 8 above, when observed relatively on the same boxplot, as shown in Figure 9, these differences are not overwhelming. The ten least represented 3-mers in A.aceti, are similarly very infrequent in E.coli, and similarly the ten most-represented 3-mers in A.aceti, are also frequently observed in E.coli. 

Overall, this analysis shows us that even though there is sequence diversity between these two bacteria species, they are nevertheless phylogenetically close.





```{r Recording Session Info}
sessionInfo()
```