---
title: "SLE777 Bioinformatics Assessment 4"
author: "Renee Carroll, s225723235"
date: "2025-10-5"
output: 
  html_document: 
    toc: true
    fig_width: 7
    fig_height: 7
theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("car")
#install.packages("seqinr")
#install.packages("R.utils")
#install.packages("ggplot2")
#install.packages("ggpubr")
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(seqinr))
suppressPackageStartupMessages(library(R.utils))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggpubr))
```

# Part 1
## Analysis of RNA-seq Expression Data

#### 1.1. Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes.

To begin the analysis, the RNA-seq expression data file must first be read into R. 

The following code will download the required data file, and will read it into R. 
The code instructs that the first row of data contains the column headings, and the first column is to be used as row names. 
It also specifies that the column separation is by tabs, and that all text columns should stay as character data, and not be converted to categorical variables.

```{r download files from GitHub, and show first 6 genes}
URL1 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
download.file(URL1, destfile="gene_expression.tsv")
gene_expression <- read.delim("gene_expression.tsv", header = TRUE, row.names = 1, sep = "\t", stringsAsFactors = FALSE)
```

The gene expression data for the first six genes for each of the three samples of interest is shown in the table below. Using the 'knitr::kable()' code ensures the document output is in a nice table format, with caption included.

```{r first 6 genes}
knitr::kable(head(gene_expression),
             caption = "Table 1. Gene expression levles for the first six genes")
```

#### 1.2. Make a new column which is the mean of the other columns. Show a table of values for the first six genes.

A new column can easily be created using the $ code. The new column is added to the end of the data frame.
To this new column, we can save the row mean calculation of the expression values from the three samples, for each gene.
The head() command displays a table of the first 6 genes from the now updated data frame.

```{r mean saved as new column}
gene_expression$mean_expression <- rowMeans(gene_expression)
knitr::kable(head(gene_expression),
             caption = "Table 2. Gene expression levles for the first six genes, including mean values")
```

#### 1.3. List the 10 genes with the highest mean expression

In order to determine the highest mean expression, the data first must be put in order according to the column that contains the mean expression values.
By including the - sign, it is ensured that the data is ordered from largest to smallest, and that is is ordering based on the column 'mean_expression'.
The comma followed by the space prior to the closing square bracket instructs that all the columns are to be kept.

With the data now ordered, the top 10 rows containing the genes with the highest mean expression can be subsetted into a new data frame.
These genes can then be listed by retrieving the row names of this subsetted data frame, in line with the text, using 'r (code)' inside the specialised R inverted commas.

```{r 10 genes with highest mean}
ordered_expression <- gene_expression[order(-gene_expression$mean_expression),]
genes_highest_expression <- head(ordered_expression, 10)
```

The 10 genes with the highest mean expression are:

1. `r rownames(genes_highest_expression)[1]`

2. `r rownames(genes_highest_expression)[2]`

3. `r rownames(genes_highest_expression)[3]`

4. `r rownames(genes_highest_expression)[4]`

5. `r rownames(genes_highest_expression)[5]`

6. `r rownames(genes_highest_expression)[6]`

7. `r rownames(genes_highest_expression)[7]`

8. `r rownames(genes_highest_expression)[8]`

9. `r rownames(genes_highest_expression)[9]`

10. `r rownames(genes_highest_expression)[10]`


#### 1.4. Determine the number of genes with a mean < 10

In order to determine the number of genes with mean expression <10, a subset is created from the main data frame with 'mean_expression<10' as the criteria for subsetting.
The total number of genes in this subset can therefore be determined by retrieving the number of rows of this new subset data frame, and the value displayed in line with text using 'r (code)' with the specialised inverted commas.

```{r number of genes with low expression < 10}
low_exp_genes <- subset(gene_expression, mean_expression<10)
```

The number of genes with a mean expression <10 is `r nrow(low_exp_genes)`.

#### 1.5. Make a histogram plot of the mean values and include it into your report.

Histograms area useful way to visualize expression data. 

Before we construct the graph, it is beneficial to first remove genes from the analysis where zero expression was detected in these samples. 
In this way, we can better focus on the expression data from the genes that are expressed in these samples.

Next, the histogram is created, with the column containing the mean expression values is selected from the appropriate data frame using the $ code.
Additional bins are included to aid in visualising the data spread. Titles for the histogram and X axis, plus the colour, are also specified.

```{r histogram of expression values, fig.cap="*Figure 1. Mean global gene expression from three samples of interest, showing severe right skewing in the data, indicating a large proportion of genes with low expression levels.*"}
gene_exp_greaterthan0 <- subset(gene_expression, mean_expression>0)
hist(gene_exp_greaterthan0$mean_expression, breaks=100, main = "Mean Gene Expression", xlab = "Mean Gene Expression", col = "dodgerblue2")
```

From this histogram, we can see that the vast majority of genes have very low expression. This severe right skewing of the data makes it difficult to glean useful information from this graph, as the large number of lowly expression genes dominates.

Data with a distribution such as this is better visualized on a logarithmic scale.

As there would be genes in this data with mean expression levels below 1, we can add 1 to all mean expression values, to ensure that the log of the values stays positive.

Here is the resulting histogram.

```{r histogram of log mean values, fig.cap="*Figure 2. Mean global gene expression from three samples of interest, on a logarithmic scale.*"}
log_gene_exp_greaterthan0 <- log(gene_exp_greaterthan0+1)
hist(log_gene_exp_greaterthan0$mean_expression, breaks=50, main = "Mean Gene Expression (log)", xlab = "Log Mean Gene Expression", col = "dodgerblue2")
```

# Part 1
## Tree Circumference Data Analysis

#### 1.6. Import this csv file into an R object. What are the column names?

The file “growth_data.csv” contains measurements for the circumferences of trees growing at two sites, a control site and a treatment site, which were planted 20 years ago.

To begin the analysis, the tree circumference data file must first be read into R. 

The following code will download the required data file, and will read it into R. 
As the file is a .csv file, and as the R code read.csv() automatically treats the first row as headings, we don't need to specify additional information. 

```{r Import csv file}
URL2 = "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
download.file(URL2, destfile="growth_data.csv")
growth_data <- read.csv("growth_data.csv") # read.csv automatically treats 
```

The column names of this data set are:

1. `r colnames(growth_data)[1]`

2. `r colnames(growth_data)[2]`

3. `r colnames(growth_data)[3]`

4. `r colnames(growth_data)[4]`

5. `r colnames(growth_data)[5]`

6. `r colnames(growth_data)[6]`

#### 1.7. Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.

To show these important data statistics, we can create a new data frame and specify the data that should populate it.
Here, the new data frame is called 'girth_stats_df'. The first column is given the name 'Location', and contains a character vector providing the names of the two sites.
Columns 2 to 5 are similarly provided first with a name, then the vector of data that is to populate those columns. 
The subset code used here requires us to specify the data frame, then the variable to subset by, which in this case is each of the locations in the Site column, followed by the column of data that's required to be copied into the new data frame, in this case, the various circumference measurements. The subset values are aggregated using the mean() or sd() function, and saved to the appropriate column. 

In this way, the data frame is built with the required information. 

For ease of data viewing, all values contained in columns 2 to 5 have been rounded to 2 decimal places.

```{r Create new data frame for the mean and sd}
girth_stats_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  Mean_2005_cm = c(mean(subset(growth_data, Site == "northeast")$Circumf_2005_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  SD_2005_cm = c(sd(subset(growth_data, Site == "northeast")$Circumf_2005_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2005_cm)),
  Mean_2020_cm = c(mean(subset(growth_data, Site == "northeast")$Circumf_2020_cm), mean(subset(growth_data, Site == "southwest")$Circumf_2020_cm)),
  SD_2020_cm = c(sd(subset(growth_data, Site == "northeast")$Circumf_2020_cm), sd(subset(growth_data, Site == "southwest")$Circumf_2020_cm))
)
girth_stats_df[,2:5] <- round(girth_stats_df[,2:5], digits = 2)
```

The following table shows the mean and standard deviation of tree circumference at the start and end of the study at both sites

``` {r Print the new data frame of mean and sd}
knitr::kable(girth_stats_df,
             caption = "Table 3. Girth statistics at the start and end of study")
```

#### 1.8. Make a box plot of tree circumference at the start and end of the study at both sites.

To create the boxplot, we'll again use the subset code to select the data we want plotted in the boxplot, from the original growth data frame. 

```{r boxplot of tree circumference at the start and end of the study at both sites, fig.cap="*Figure 3. Boxplot of tree girth measurements (in cm) taken in 2005 and 2020 at the Norheast and Southwest study sites. The median value is shown by the black horizontal central line. The interquartile range, comprised of the upper quartile above the median, and the lower quartile below the median, is represented by the coloured box. The maximum and minimum values are shown by the ends of the whiskers.*"}
boxplot(
  subset(growth_data, Site == "northeast")$Circumf_2005_cm, 
  subset(growth_data, Site == "northeast")$Circumf_2020_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2005_cm, 
  subset(growth_data, Site == "southwest")$Circumf_2020_cm, 
  names = c("Northeast 2005", "Northeast 2020", "Southwest 2005", "Southwest 2020"), 
  col = c("forestgreen"),
  ylab = "Circumference (cm)", 
  main = "Increase in tree girth between 2005 and 2020, at each study site")
```

#### 1.9. Calculate the mean growth over the last 10 years at each site.

To calculate and display the mean growth from 2010 to 2020, use the $ code to add a column to the original data frame to contain this information. 
To this column, assign the difference in circumference measurements from 2010 to 2020.
Next, create a new data frame, and save into this the site names, and the calculated mean of the difference in measurements from the main data frame, using the subset() code to select the appropriate data as per each location. Including the code 'check.names=FALSE' ensures that the column names appear as intended.

```{r mean growth over the last 10 years}
# Add a column to the existing data frame, to contain the calculation of 2020 measurements minus the 2010 measurements.
growth_data$growth_2010_2020 <- growth_data$Circumf_2020_cm-growth_data$Circumf_2010_cm 

# Create a new data frame to present the mean of the 10 year growth data, for each location.
mean_10yr_growth_df <- data.frame(
  Location = c("Northeast", "Southwest"),
  "Mean 10 year growth (cm)" = c(mean(subset(growth_data, Site == "northeast")$growth_2010_2020), mean(subset(growth_data, Site == "southwest")$growth_2010_2020)),
  check.names = FALSE 
)
```

Here is the mean growth, measured in cm, over the last 10 years at each of the sites.

```{r table showing mean growth over 10 years}
knitr::kable(mean_10yr_growth_df,
             caption = "Table 4. Mean growth over last 10 years")
```

#### 1.10. Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.

Prior to conducting a t-test, it is important to fist check for equal variance in the data.
Equal variance can be calculated using leveneTest() code as part of the "car" package.

``` {r}
leveneTest(growth_2010_2020 ~ Site, data = growth_data)
```
As the Levene's Test returned a P-value of 0.006116, it indicates the variances in this data are not equal, and therefore the appropriate t-test to use is a Welch's t-test, as it adjusts for unequal variances.

The output for Welch's t-test is below:

```{r t.test)}
t.test(growth_2010_2020 ~ Site, data = growth_data, var.equal = FALSE) # var.equal = FALSE isn't required here, as it is the default for t.test() in R.
```
This t-test provided a p-value of 0.06229, indicating the difference in 10 year growth between the two sites is not statistically significant, however as this value is only slightly above the cutoff of 0.05, is is close to being significant.

The recommendation is to keep monitoring tree growth at the two sites, and repeat the analysis in the future.


# Part 2
## Examining biological sequence diversity between E.coli and A.aceti

Here in this report, we will perform a comparative analysis on the coding sequences of two bacteria, Escherichia coli (E.coli) and Acetobacter aceti (A.aceti).

The analysis utilizes two packages installed and loaded earlier; the Seqinr package, to process and analyse sequence data, and the R.utils package, which is a general utilities package enabling files to be unzipped. 

#### 2.1.1. Download the whole set of coding DNA sequences for E. coli and your organism of interest.

The coding DNA sequence files (cds files) can be downloaded from Ensembl's database, using the respective URLs in the code chunk below. 
Once downloaded and saved, the FASTA file can then be unzipped using the gunzip() code from the "R.utils" package.

Both FASTA files are read into R using the read.fasta() code as part of the "seqinr" package.

```{r Download E.coli coding DNA sequences}
# Download the E.coli cds, and unzip the file.
#URL3 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_0_collection/escherichia_coli_str_k_12_substr_mg1655_gca_000005845/cds/Escherichia_coli_str_k_12_substr_mg1655_gca_000005845.ASM584v2.cds.all.fa.gz"
#download.file(URL3, destfile="ecoli_cds.fa.gz")
#gunzip("ecoli_cds.fa.gz")

# Download the cds for my organism of interest, Acetobacter aceti GCA_002005445, and unzip the file.
#URL4 = "https://ftp.ensemblgenomes.ebi.ac.uk/pub/bacteria/release-62/fasta/bacteria_80_collection/acetobacter_aceti_gca_002005445/cds/Acetobacter_aceti_gca_002005445.ASM200544v1.cds.all.fa.gz"
#download.file(URL4, destfile="aaceti_cds.fa.gz")
#gunzip("aaceti_cds.fa.gz")

# Read in the two sequence files.
ecoli_cds <- seqinr::read.fasta("ecoli_cds.fa")
aaceti_cds <- seqinr::read.fasta("aaceti_cds.fa")
```

#### 2.1.2. How many coding sequences are present in these organisms? Present this in the form of a table. 

Within the FASTA files, each line contains the coding DNA sequence for a different gene. Therefore, to number of coding seuqences present in each organism is given by the number of rows in the FASTA file. In this context, the length() code provides us with this answer. By creating a new data frame with this data, we can then easily present it in the form of a table.

```{r Table of the number of coding sequences present in these organisms}
cds_df <- data.frame(
  Organism = c("E.coli", "A.aceti"),
  "Number of Coding Sequences" = c(length(ecoli_cds), length(aaceti_cds)),
  check.names = FALSE 
  )
knitr::kable(cds_df,
             caption = "Table 5. Number of coding sequences present in E.coli and A.aceti")
```

#### 2.1.3. Describe any differences between the two organisms. 

(Note: The following sentences contain in line code)

The organism E.coli has `r (length(ecoli_cds)-length(aaceti_cds))` additional genes in its coding DNA, which is an increase of `r round((length(ecoli_cds)-length(aaceti_cds))/length(aaceti_cds)*100, 1)`%. 

This data suggests that E.coli is a more complex organism than A.aceti, as a greater number of genes would result in a greater number of proteins, and hence additional or more complex biological functionality.

#### 2. How much coding DNA is there in total for these two organisms? Present this in the form of a table.

In order to determine the total amount of coding DNA for each organism, we can perform the 'summary()' code to show the length of each coding sequence (presented in the first column). Then, we can take the values from this first column and save this information as a new number vector. Taking the sum of these new vectors gives us the total amount of coding DNA for each organism, which can then be saved as a data frame and presented in the form of a table, as shown below. 

```{r Table of the total amount of coding DNA for the two organisms}
length_ecoli_genes <- as.numeric(summary(ecoli_cds)[,1])
length_aaceti_genes <- as.numeric(summary(aaceti_cds)[,1])
cds_total_df <- data.frame(
  Organism = c("E.coli", "A.aceti"), 
  "Total Amount of Coding Sequence" = c(sum(length_ecoli_genes), sum(length_aaceti_genes)),
  check.names = FALSE 
) # Including the check.names = FALSE code so that R doesn't add '.' in between the words of the Column names.
knitr::kable(cds_total_df,
             caption = "Table 6. Total quantity of coding DNA in E.coli and A.aceti")
```

#### Describe any differences between the two organisms. 

(Note: The following sentences contain in line code)

In line with the previous observation that E.coli contains `r round((length(ecoli_cds)-length(aaceti_cds))/length(aaceti_cds)*100, 1)`% more genes than A.aceti, we can further determine that E.coli has a proportinal increase in the total amount of coding sequence as well, namely, that E.coli contains `r  round((sum(length_ecoli_genes)-sum(length_aaceti_genes))/sum(length_aaceti_genes)*100, 1)`% more coding DNA in total. This further supports the hypothesis that E.coli is a more complex organism than A.aceti.

### 3. Calculate the length of all coding sequences in these two organisms. Make a boxplot of coding sequence length in these organisms. 

Using the vectors of cDNA lengths just created, the boxplot can be easily generated. 

```{r Boxplot of coding sequence length in these organisms, fig.cap="*Figure 4. Distributions of coding sequence lengths between two organisms, E.coli (red) and A.aceti (blue). The median values are shown by the black horizontal central lines. The interquartile range, comprised of the upper quartile above the median, and the lower quartile below the median, are represented by the coloured boxes. The maximum and minimum values are shown by the ends of the whiskers, with outlying values shown by points above the whisker border line.*"}
boxplot(length_ecoli_genes, length_aaceti_genes, names = c("E.coli", "A.aceti"), main = "Coding DNA Sequence Lengths in E.coli and A.aceti", ylab="Length of cDNA (bp)", col=c("red", "blue"))
```

### What is the mean and median coding sequence length of these two organisms? Describe any differences between the two organisms. 

(Note: The following sentences contain in line code)

The average length of a coding sequence in E.coli is `r round(mean(length_ecoli_genes), 0)` bp long, and in A.aceti is `r round(mean(length_aaceti_genes), 0)` bp long. The median coding sequence length in E.coli is `r median(length_ecoli_genes)`, and in A.aceti is `r median(length_aaceti_genes)`. 

Together with the boxplot shows in Figure 4 above, this data reveals that the transcriptomes of E.coli and A.aceti are comparable in terms of gene length. In both organisms, the vast majority of coding sequences are around the 800-900 bp length, with both organisms having a smaller number of very large genes, in the approximate range of 2500-8000 bp long, as shown by the data points above the plots' upper whisker lines. The presence of these large genes are the reason the calculated average length is larger than the calculated median length, in both organisms. 

### 4. Calculate the frequency of DNA bases in the total coding sequences for both organisms. Create a bar plot for nucleotide frequency. 

Using the 'unlist()' command, the DNA sequences from each organisms coding DNA sequence (cds) data frame can be saved as a separate vector. The 'count()' command can then be used on this vector, with the instruction to count the number of nucleotides present in the vector in combinations of 1's (ie. single nucleotides, a, c, g, and t), saving this result as another object.

The proportion of each nucleotide in each organism's total coding sequence can then be calculated, with the results graphed in a grouped bar graph, using the packages "ggplot2" and "ggpubr" (see Figure 5).

```{r Nucleotide composition and frequencies in E.coli and A.aceti}

# Determining DNA composition for each organism
dna_ecoli <- unlist(ecoli_cds) 
dna_composition_ecoli <- count(dna_ecoli,1)

dna_aaceti <- unlist(aaceti_cds) 
dna_composition_aaceti <- count(dna_aaceti,1) 

# Determining DNA proportion for each organism
dna_proportion_ecoli <- dna_composition_ecoli/sum(dna_composition_ecoli)
dna_proportion_ecoli <- round(dna_proportion_ecoli, digits = 3) # digits=3 means that the result is rounded to 3 decimal places.

dna_proportion_aaceti <- dna_composition_aaceti/sum(dna_composition_aaceti)
dna_proportion_aaceti <- round(dna_proportion_aaceti, digits = 3)
```

``` {r Creating the bar chart for nucleotide frequencies using ggplot2, fig.cap="*Figure 5. Grouped bar chat showing the frequencies of nucleotide composition in the two orgfanisms, E.coli and A.aceti.*"}
# Create data frames from the nucleotide proportions, and add the column for Organism
dna_proportion_ecoli_df <- as.data.frame(dna_proportion_ecoli)
dna_proportion_ecoli_df$Organism <- c("E.coli")

dna_proportion_aaceti_df <- as.data.frame(dna_proportion_aaceti)
dna_proportion_aaceti_df$Organism <- c("A.aceti")

# Merge the two data frames together, to create one data frame for the bar plot
proportion_df <- rbind(dna_proportion_ecoli_df, dna_proportion_aaceti_df)
names(proportion_df) <- c("Nucleotide", "Proportion", "Organism") 

# Define the variables for the grouped bar chart
Organism <- proportion_df$Organism
Nucleotide <- proportion_df$Nucleotide
Proportion <- proportion_df$Proportion

# Create the grouped bar chart. 
# The code specifies the data frame to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A blue colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Blues".
ggbarplot(data=proportion_df, x="Nucleotide", y="Proportion", fill="Organism", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Blues") 
```

### Describe any differences between the two organisms.

While nucleotide composition is comparable between E.coli and A.aceti, it is observed that the coding genome of the organism of interest, A.aceti, contains an decreased proportion of adenine (a) and thymine (t), and an increased proportion of cytosine (c) and guanine (g), as compared to the coding genome of E.coli. 

### 4. Calculate the frequency of amino acids in the total protein sequences for both organisms. Create a bar plot for amino acid frequency. 

In order to analyse the protein sequences in these organisms, the coding sequences must first be translated.
Here, the lapply function performs a loop, applying the 'translate' code to all elements of each organisms's 'cds' data, and the output is saved as new object. 

Next, the 'unlist()' command is again used to extract the amino acid sequences from the newly created data objects, and saving the total protein sequence as a new single vector. 

We then use the count() command again, but we first must define the list of possible amino acids which the count() command is to reference. To do this, we can determine the unique elements present in one of the organism's proteome vectors, remove all stop codons, and save it as a new vector 'aa', as outlined in the code chunk below. When the count() command is used here, we then define the alphabet to be the newly created 'aa' vector, and the wordsize to be 1 (ie. single amino acids).

The proportion of each amino acid in each organism's total protein sequence can then be calculated, with the results graphed in a grouped bar graph, using the packages "ggplot2" and "ggpubr" (see Figure 6).

```{r Determining the frequency of amino acids in the total protein sequences of each organism}
# Translate the cDNA for each organism
ecoli_prot <- lapply(ecoli_cds, translate)
aaceti_prot <- lapply(aaceti_cds, translate)

# Unlist the protein sequences and saving then as a single vector
ecoli_prot_total <- unlist(ecoli_prot)
aaceti_prot_total <- unlist(aaceti_prot) 

# Creates the list of possible amino acids to form our alphabet to use in the 'count' code.
aa <- unique(ecoli_prot_total) 

# Removes the stop codon asterisks.
aa <- aa[aa != "*"] 

# Count the number of each amino acid in each organisms proteome vector, in combinations of 1's (ie. single amino acids), using the aa alphabet previously determined, and save it as a new object. 
aa_composition_ecoli <- count(ecoli_prot_total, wordsize=1, alphabet=aa) 
aa_composition_aaceti <- count(aaceti_prot_total, wordsize=1, alphabet=aa)

# Calculate the proportion of each amino acid in each organisms' proteome
aa_proportion_ecoli <- aa_composition_ecoli/sum(aa_composition_ecoli)
aa_proportion_ecoli <- round(aa_proportion_ecoli, digits = 3)
aa_proportion_aaceti <- aa_composition_aaceti/sum(aa_composition_aaceti)
aa_proportion_aaceti <- round(aa_proportion_aaceti, digits = 3)
```


###### RESUME FROM HERE #########


```{r Bar chat for amino acid frequencies}
#Create data frames from the aa proportions, which are currently in table format, and add the column for Organism
aa_proportion_ecoli_df <- as.data.frame(aa_proportion_ecoli)
aa_proportion_ecoli_df$Organism <- c("E.coli")
aa_proportion_aaceti_df <- as.data.frame(aa_proportion_aaceti)
aa_proportion_aaceti_df$Organism <- c("A.aceti")

#Merge the two data frames together, to be one data frame
aa_proportion_df <- rbind(aa_proportion_ecoli_df, aa_proportion_aaceti_df)
names(aa_proportion_df) <- c("Amino_Acids", "Proportion_aa", "Organism_aa") 
knitr::kable(aa_proportion_df,
             caption = "Table 7. ____")
```

Making the bar graphs
```{r bar chart for amino acid frequencies}
#Defining the variables for the grouped bar chart
Organism_aa <- aa_proportion_df$Organism_aa
Amino_Acids <- aa_proportion_df$Amino_Acids
Proportion_aa <- aa_proportion_df$Proportion_aa

#Create the grouped bar chart
#The code specifies the dataframe to use, and then which variables to plot where. It defines the position is side-by-side with the code 'position=position_dodge', and ensures there is a space between the columns with the code '(0.8)'. A green colour palette was chosen for the graph with the code scale_fill_brewer(palette = "Greens".
ggbarplot(data=aa_proportion_df, x="Amino_Acids", y="Proportion_aa", fill="Organism_aa", position=position_dodge (0.8))+
  scale_fill_brewer(palette = "Greens") 
```

### Describe any differences between the two organisms. 

















### 5. Create a codon usage table and quantify the codon usage bias among all coding sequences. Describe any differences between the two organisms with respect to their codon usage bias. Provide charts to support your observations.

The RSCU stands for relative synonymous codon usage, which is a measure as to how preferred a codon of a particular amino acid is compared to the other codons which also encode it. A value of 1 indicating that the codon is used as expected. Values above 1 suggest that the codon is used more frequently than expected, while values below 1 indicate lower usage. For instance, codons like "aaa" and "att" show higher RSCU values, indicating they are used more often, whereas codons like "tga" and "tag" have lower values, suggesting less frequent use.

```{r Creating the codon usage table FIRST ATTEMPT, BUT THE ROWS ARE JUST ADDED TO EACH OTHER. SHOULD I HAVE INSTEAD ADDED JUST THE COLUMN DATA?}
# Create the codon usage data frame for E.coli, showing RSCU values. 
codons_ecoli_df <- uco(dna_ecoli,index="rscu",as.data.frame=TRUE)
print(codons_ecoli_df)

# Create the codon usage data frame for A.aceti, showing RSCU values
codons_aaceti_df <- uco(dna_aaceti,index="rscu",as.data.frame=TRUE)
print(codons_aaceti_df)
```

#### IF I HAVE TIME IT WOULD BE GOOD TO PUT THIS IN ONE TABLE, THOUGH IM NOT SURE IF THAT ONE TABLE WOULD BE EASY TO GRAPH TOGETHER OR NOT.
```{r ATTEMPT TO ADD THESE INTO ONE TABLE}
# Set the column names to include which organism it is

# OR CREATE A NEW DATAFRAME, WITH SELECTED FEATURES OF THE OTHER TWO, TO MAKE A DATAFRAME THAT HAS JUST THE AA, CODON, AND RSCU VALUE.

# Merge the rows required
```

```{r Grouped bar chart for RSCU values, minus 1}
# Add a column for organism name
codons_ecoli_df$Organism <- c("E.coli")

# Add a column for organism name
codons_aaceti_df$Organism <- c("A.aceti")

# Merge these dataframes into one
column_names <- colnames(codons_ecoli_df)
codons_df <- rbind(codons_ecoli_df, codons_aaceti_df)
names(codons_df) <- column_names
print(codons_df)

# Create a new column that calculates the RSCU value minus 1, such that values are either above or below a zero line.
codons_df$"RSCU-1" <- codons_df$RSCU-1

# Create a new column that contains the amino acid name plus it's codon
codons_df$"AA_and_codon" <- paste(codons_df$AA, ":", codons_df$codon)

# Put the data in alphabetical order, according to the amino acid 3 letter code, and print the resulting table
ordered_codons_df <- codons_df[order(codons_df$AA), ]
print(ordered_codons_df)
```

```{r Grouped barplot showing RSCU-1 values for E.coli and A.aceti}
# In the theme() component, I've set the column text size to 7, have adjusted the angle to 90 degrees,have justified the lables to be straight against the x-axis, and have shifted them left such that they line up under the tick lines. 
ggbarplot(data=ordered_codons_df, x="AA_and_codon", y="RSCU-1", fill="Organism", position=position_dodge (0.8)) +
  scale_fill_brewer(palette = "Oranges") +
  ggtitle("Relative Synonymous Codon Usage between E.coli and A.aceti") +
  xlab("Amino Acid : Codon") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust=1, vjust=0.5))
```




### 6. In the organism of interest, identify 10 protein sequence k-mers of length 3-5 which are the most over- and under-represented k-mers in your organism of interest. Are these k-mers also over- and under-represented in E. coli to a similar extent? Provide plots to support your observations. Why do you think these sequences are present at different levels in the genomes of these organisms?

```{r Extracting K-mers frequency for sizes 3, 4 and 5, and saving as separate data frames}
# Determine all K-mers of size 3, and their frequency of use
aaceti_kmer3 <- data.frame(count(aaceti_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer3) <- c("Kmer", "Frequency")

# Order this data, lowest to highest, then extract the top 10 and save in a new data frame called 'aaceti_kmer3_bottom10'
aaceti_kmer3 <- aaceti_kmer3[order(aaceti_kmer3$Frequency), ] 
aaceti_kmer3_bottom10 <- head(aaceti_kmer3, 10)
print(aaceti_kmer3_bottom10)

# Order this data, from highest use to lowest, then extract the top 10 and save in a new data frame called 'aaceti_kmer3_top10'
aaceti_kmer3 <- aaceti_kmer3[order(-aaceti_kmer3$Frequency), ] 
aaceti_kmer3_top10 <- head(aaceti_kmer3, 10)
print(aaceti_kmer3_top10)

# Repeat this process for all K-mers of size 4 amino acids
aaceti_kmer4 <- data.frame(count(aaceti_prot_total, wordsize=4, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer4) <- c("Kmer", "Frequency")

aaceti_kmer4 <- aaceti_kmer4[order(aaceti_kmer4$Frequency), ] # Orders by frequency, smallest to largest
aaceti_kmer4_bottom10 <- head(aaceti_kmer4, 10)
print(aaceti_kmer4_bottom10)

aaceti_kmer4 <- aaceti_kmer4[order(-aaceti_kmer4$Frequency), ] # Inclusion of - means this is now ordered by frequency, largest to smallest
aaceti_kmer4_top10 <- head(aaceti_kmer4, 10)
print(aaceti_kmer4_top10)

# Repeat this process for all K-mers of size 5 amino acids
aaceti_kmer5 <- data.frame(count(aaceti_prot_total, wordsize=5, alphabet=aa, freq=TRUE))
colnames(aaceti_kmer5) <- c("Kmer", "Frequency")

aaceti_kmer5 <- aaceti_kmer5[order(aaceti_kmer5$Frequency), ] # Orders by frequency, smallest to largest
aaceti_kmer5_bottom10 <- head(aaceti_kmer5, 10)
print(aaceti_kmer5_bottom10)

aaceti_kmer5 <- aaceti_kmer5[order(-aaceti_kmer5$Frequency), ] # Inclusion of - means this is now ordered by frequency, largest to smallest
aaceti_kmer5_top10 <- head(aaceti_kmer5, 10)
print(aaceti_kmer5_top10)
```

Given that there are K-mers of size 4 and 5 not used in A.aceti, whereas K-mers of size 3 are all present in the A.aceti proteome, further comparisons between A.aceti and E.coli will focus just on the 10 most over- and under-represented k-mers of size 3.

```{r Extracting lists of the 10 most over- and under- represented 3 K-mers in A.aceti}
# Extract list of bottom 10 3-K-mers
Bottom10_kmer3 <- as.vector(aaceti_kmer3_bottom10$Kmer)
print(Bottom10_kmer3)

# Extract list of top 10 3-K-mers
Top10_kmer3 <- as.vector(aaceti_kmer3_top10$Kmer)
print(Top10_kmer3)
```

```{r Extracting frequencies of K-mers of size 3 in E.coli}
# Determine all K-mers of size 3, and their frequency of use, in E.coli
ecoli_kmer3 <- data.frame(count(ecoli_prot_total, wordsize=3, alphabet=aa, freq=TRUE))
colnames(ecoli_kmer3) <- c("Kmer", "Frequency")
```

```{r}
# Subset the E.coli list, using the top 10 A.aceti list, and save as a new df
ecoli_kmer3_topsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Top10_kmer3, ]

# Add a column for Organism, and note is as E.coli
ecoli_kmer3_topsubset$Organism <- c("E.coli")
print(ecoli_kmer3_topsubset)

# Subset the E.coli list, using the bottom 10 A.aceti list, and save as a new df
ecoli_kmer3_bottomsubset <- ecoli_kmer3[ecoli_kmer3$Kmer %in% Bottom10_kmer3, ]

# Add a column for Organism, and note is as E.coli
ecoli_kmer3_bottomsubset$Organism <- c("E.coli")
print(ecoli_kmer3_bottomsubset)
```

```{r}
#Prepare the A.aceti K-mer 3 data frames, by adding a column for organism
aaceti_kmer3_bottom10$Organism <- c("A.aceti")
aaceti_kmer3_top10$Organism <- c("A.aceti")

# Merge the two organisms' data frames that contain the frequency data for the top 10 3-K-mers from A.aceti
column_names_kmer3 <- colnames(aaceti_kmer3_bottom10)

Top_kmer3_df <- rbind(aaceti_kmer3_top10, ecoli_kmer3_topsubset)
names(Top_kmer3_df) <- column_names_kmer3
print(Top_kmer3_df)

# Merge the two organisms' data frames that contain the frequency data for the bottom 10 3-K-mers from A.aceti
Bottom_kmer3_df <- rbind(aaceti_kmer3_bottom10, ecoli_kmer3_bottomsubset)
names(Bottom_kmer3_df) <- column_names_kmer3
print(Bottom_kmer3_df)
```

```{r Boxplot comparison of A.aceti top 10 k-mers}
ggbarplot(data=Top_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  ggtitle("10 most over-represented 3 amino acid k-mers in A.aceti") +
  scale_fill_brewer(palette = "Greens") 
```

```{r Boxplot comparison of A.aceti bottom 10 k-mers}
ggbarplot(data=Bottom_kmer3_df, x="Kmer", y="Frequency", fill="Organism", position=position_dodge (0.8))+
  ggtitle("10 most under-represented 3 amino acid k-mers in A.aceti") +
  scale_fill_brewer(palette = "Reds") 
```



```{r Recording Session Info}
sessionInfo()
```